Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-11 paper/introduction.tex
Updates to introduction

Consider the example shown in Fig.~\ref{fig:motivating-eg}.
This code represents the logic for a streaming query operator (also
known as a continuous query). The operator receives a continuous
stream of input messages, which are processed via a user-defined
function, and batched into time-windows associated with a logical
timestamp. Messages may be delivered out-of-order and multiple
messages with the same timestamp may be delivered. A timing-message,
periodically indicates that no more input messages with a timestamp
less than or equal to a given value $t$ will be subsequently
delivered. At this point, the operator can create the output message
corresponding to time $t$ and deliver it to its successors.

@@

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer.  In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, the garbage collection overheads become
significant.
% as the collector scans these large data-structures
However, these data-structures have a well-defined lifetime, which can
be exploited to achieve better performance 
% using alternative memory management techniques.

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of one or more contiguous
range of memory locations, permitting a region to grow) that is used to realize
a data-structure with a well-defined lifetime. 
Once a region is created, individual objects may be dynamically allocated within
the region over time. Finally, the entire region is typically freed in one shot.

Region-based memory management, both manual as well as automatic, has been known for
a long time. Manual region-based memory management suffers from the usual challenges,
namely the potential for dangling references and lack of memory safety. Automatic
region-based memory management systems guarantee memory safety, but impose various restrictions.
Tofte and Talpin's system, for example, requires that the set of all regions (existing
at a point in time) forms a stack. Specifically, this requires that the lifetimes of
all regions must be well-nested: it is not possible to have two regions whose lifetimes
overlap, with neither one's lifetime contained within the other.
Unfortunately, the data-structures in the above example do not satisfy this restriction.
We refer to such regions as \emph{dynamic} regions.

@@

We use \emph{static regions} (regions with lexically-scoped lifetimes), as well as the stack,
for such temporary pointers (to objects inside a dynamic region) and use the following protocol
to ensure safety of such pointers.
To work with a dynamic region $D$, we must first \emph{open} region $D$,
to indicate that $D$ should not be freed during this period.
When the processing of the region is complete, we close the region $D$,
to indicate that it is safe to free $D$.
Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,
but uses lexical scoping to ensure that such pointers are no longer live when the region is closed.
Thus, such pointers may be either stack-allocated or reside in a region whose lifetime is statically
guaranteed to be contained within the interval when $D$ is open.
As a consequence, the system ensures the strong invariant that for a \emph{closed dynamic} region $D$
no pointer from outside $D$ points to an internal object of $D$.

>>>>

Consider the example shown in Fig.~\ref{fig:motivating-eg}.
This code represents the logic for a streaming query operator.
The operator receives a stream of input messages, each associated
with a time (window) $t$, processed by method \texttt{onReceive}.
Each input message contains a list of inputs, each of which is processed
by applying a user-defined function to create a corresponding output.
The operator may receive multiple messages with the same timestamp
(and messages with different timestamps may be delivered out of order).
A timing-message (an invocation of method \texttt{OnNotify}) indicates that
no more input messages with a timestamp $t$ will be subsequently delivered.
At this point, the operator completes the processing for time window $t$ and sends
a corresponding output message to its successor.

@@

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer.  In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, the garbage collection overhead becomes
significant.
% as the collector scans these large data-structures
However, these data-structures have a well-defined lifetime, which can
be exploited to achieve better performance.
% using alternative memory management techniques.

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of one or more contiguous
range of memory locations, permitting a region to grow) that is used to realize
a data-structure with a well-defined lifetime. 
Once a region is created, individual objects may be dynamically allocated within
the region over time. Finally, the entire region is typically freed in one shot.
In this example, the output to be constructed for each time window $t$ can be
implemented using a separate region.

Region-based memory management, both manual as well as automatic, has been known for
a long time. Manual region-based memory management suffers from the usual challenges,
namely the potential for dangling references and lack of memory safety. Automatic
region-based memory management systems guarantee memory safety, but impose various restrictions.
Tofte and Talpin's system~\cite{tofte94,tofte97} , for example, uses lexically scoped regions.
At runtime, the set of all regions (existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two regions whose lifetimes
overlap, with neither one's lifetime contained within the other.
Unfortunately, the data-structures in the above example do not satisfy this restriction
(as the output messages for multiple time windows may be simultaneously live, without
any containment relation between their lifetimes).
We refer to regions with lexically scoped lifetimes as \emph{stack regions} and to regions 
that do not have such a lexically scoped region as \emph{dynamic} regions.

@@

We use \emph{stack regions} (regions with lexically-scoped lifetimes), as well as the stack,
for such temporary pointers (to objects inside a dynamic region) and use the following protocol
to ensure safety of such pointers.
To work with a dynamic region $D$, we must first \emph{open} region $D$,
to indicate that $D$ should not be freed during this period.
When the processing of the region is complete, we close the region $D$,
to indicate that it is safe to free $D$.
In general, a dynamic region may be opened and closed multiple times before it is finally freed.

Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,
but uses lexical scoping to ensure that such pointers are no longer live when the region is closed.
Thus, such pointers may be either stack-allocated or reside in a region whose lifetime is statically
guaranteed to be contained within the interval when $D$ is open.
As a consequence, the system ensures the strong invariant that for a \emph{closed dynamic} region $D$
no pointer from outside $D$ points to an internal object of $D$.
